---
title: "Preprocessing_notrim_Ella"
output: html_document
---

```{r, warning = FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
```

```{r, warning = FALSE}
#get names
var_names <- read_csv('../data/biomarker-raw.csv', 
                     col_names = FALSE, 
                     n_max = 2, 
                     col_select = -(1:2)) %>%
  t() %>%
  as_tibble() %>%
  rename(name = V1, abbreviation = V2) %>%
  na.omit()

# No trim
trim <- function(x, .at){
  x[abs(x) > .at] <- sign(x[abs(x) > .at]) * .at
  return(x)
}

# log + scale
biomarker_clean_notrim <- read_csv('../data/biomarker-raw.csv', 
         skip = 2,
         col_select = -2L,
         col_names = c('group', 
                       'empty',
                       pull(var_names, abbreviation),
                       'ados'),
         na = c('-', '')) %>%
  filter(!is.na(group)) %>%
  # log-transform + standardize only
  mutate(across(.cols = -c(group, ados), 
                ~ scale(log10(.x))[, 1])) %>%
  # reorder columns
  select(group, ados, everything())

# Save as no dataset

save(list = 'biomarker_clean_notrim', 
     file = '../data/biomarker-clean-NOTRIM.RData')
```


# Difference between 

```{r}
diff_summary <- summary(abs(biomarker_clean_notrim[,-c(1,2)] - biomarker_clean[,-c(1,2)]))
diff_summary

data_numeric <- biomarker_clean_notrim %>%
  select(-group, -ados)

outlier_count <- apply(data_numeric, 1, function(x) sum(abs(x) > 3, na.rm = TRUE))
outlier_table <- tibble(
  group = biomarker_clean_notrim$group,
  outlier_count = outlier_count
)

summary(outlier_table$outlier_count)
table(outlier_table$group)

outlier_summary <- outlier_table %>%
  group_by(group) %>%
  summarise(
    mean_outliers = mean(outlier_count),
    median_outliers = median(outlier_count),
    max_outliers = max(outlier_count)
  )

outlier_summary

# Figure x
group_means <- outlier_table %>%
  group_by(group) %>%
  summarise(
    mean_outliers = mean(outlier_count),
    se_outliers = sd(outlier_count)/sqrt(n())
  )

ggplot(group_means, aes(x = group, y = mean_outliers, fill = group)) +
  geom_col(alpha = 0.6) +
  geom_errorbar(aes(ymin = mean_outliers - se_outliers,
                    ymax = mean_outliers + se_outliers),
                width = 0.2) +
  theme_minimal() +
  labs(
    title = "Average Number of Outlying Proteins per Subject",
    x = "Group",
    y = "Mean ± SE of Outlier Counts"
  ) +
  theme(legend.position = "none")

```

# summary 

To explore the influence of trimming, we repeated preprocessing without removing extreme values (|z| ${>}$3).
For each subject, we counted the number of protein measurements whose standardized
log-transformed value exceeded ${\pm}$3.
This allowed us to identify subjects with unusually large numbers of outlying proteins.
On average, TD subjects had slightly more outlying protein values (mean = 17.6) than ASD subjects (mean = 13.3), though medians were identical (8.5).

This indicates that while a few TD participants show many extreme measurements, group-level differences are small.

Figure X shows the mean ${\pm}$SE of outlier counts per subject by group.
